A novel framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall

### 2019-07-18 paper/main.tex

%\title{How is experience transformed into memory?}
%\title{A content-based framework for capturing stimulus and memory dynamics reveals event-like structure in naturalistic episodic recall}
%\title{A content-based model characterizing/describing/representing how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall}
%\title{A content-based framework describing the temporal dynamics of experiences and memories reveals event-like structure in naturalistic episodic recall}
%\title{A content-based model of how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall}
\title{A framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall}

@@

\begin{abstract}
Our life experiences unfold over time and the temporal dynamics of their contents form unique \textit{experience trajectories}.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  We apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We find that the video and subsequent recall share an event-like structure, and that the shapes of the trajectories formed by participants' recountings were all highly similar to that of the original video. Further, the level of precision that participants' recount the events as well as how distinct their recountings are from other recall events both predict overall subsequent memory performance. Lastly, we identified a network of brain structures that are sensitive to the ``shapes'' of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences. These result highlight that our dynamic experiences are structured into events in memory, and introduce a novel framework for mapping between dynamic life experiences and their mnemonic counterparts.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Undoubtedly, using well-controlled trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our memories are continuous, rather than discrete: removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, counting the number or proportion of precise recalls is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or shape) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener.

@@

% NOTE: Can we remove this part? its captured in the abstract
% Using a series of novel memory metrics (introduced here and afforded by our computational framework), we find that the successful memory performance is related to 1) the \textit{precision} with which the participant recounts each event and 2) how \textit{distinctive} each recall event is (relative to the other recalled events). Finally, we identified a network of brain structures whose responses (as participants watched the episode) reflected the shape of the episode, and a second network whose responses reflected how participants would later recount the episode.

\section*{Results}
To characterize the shape of the \textit{Sherlock} episode participants watched and their subsequent recountings of the episode, we used a topic model~\citep{BleiEtal03} to discover the latent thematic content in the video.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents; they return as output two matrices.  The first output is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes which mix topics is reflected in each document.

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus events and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging this affordance, we introduce 2 novel metrics for quantifying naturalistic memory representations: precision and distinctiveness.  We define precision as the average correlation between each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that across participants, a higher precision score is correlated to both hand annotated memory performance (Pearson's $r(15) = 0.6, p = 0.011$) as well as the number of recall events (per participant) estimated by our model (Pearson's $r(15) = 0.64, p = 0.005$). A second novel metric we introduce here is the distinctiveness, or the uniqueness of each recall event relative to other recall events. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  Similar to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.83, p < 0.001$ and model derived memory: Pearson's $r(15) = 0.71, p = 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.

@@

The dynamic content of the video and participants' recalls is quantified in the corresponding topic proportion matrices.  However, it is difficult to gain deep insights into that content solely by examining the topic proportion matrices (e.g., Figs.~\ref{fig:model}A, D) or the corresponding correlation matrices (Figs.~\ref{fig:model}B, E, \corrmats).  To visualize the time-varying high-dimensional content in a more intuitive way~\citep{HeusEtal18a} we projected the topic proportions matrices onto a two-dimensional space using Uniform Manifold Approximation and Projection~\citep[UMAP; ][]{McInHeal18}.  In this lower-dimensional space, each point represents a single video or recall event, and the distances between the points reflect the distances between the events' associated topic vectors (Fig.~\ref{fig:trajectory}).

@@

When modeling memory experiments, often times events (or items) and their later memories are treated as binary (or categorical in the case of confidence ratings) and independent events. Our novel framework allows one to assess memory performance in a more continuous way, as well as analyze the correlational structure of each encoding event to each memory event. Further and importantly, it allows for consideration of the actual content of the experience/memories, which is not typically modeled. Leveraging this, using 2 novel memory metrics we find that the successful memory performance is related to 1) the \textit{precision} with which the participant recounts each event and 2) how \textit{distinctive} each recall event is (relative to the other recalled events). The first finding suggests to us that the accuracy of recall for \textit{any individual event} may predict the overall amount of information recovered by the participant.  The second finding suggests that remembering/describing events in a unique way (relative to other recalled events) is also related to the quantity of content recovered. Intriguingly, prior studies show that pattern separation, or the ability to discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging [Craig Stark and friends].  Future work might explore how/whether these metrics compare between cognitively impoverished groups and healthy controls.

While a large number of language models exist (e.g. WAS, LSA, word2vec, universal sentence encoder), here we use topic models for a few reasons [CITE THESE].  First, topic models capture the \textit{essence} of a text passage devoid of the specific set and order of words used.  This was an important feature of our model since different people may accurately recall a scene using very different language. Secondly, words can mean different things in different contexts (e.g. baseball bat vs. the animal bat), and topic models are robust to this since words can be part of multiple topics.  Lastly, topic models provide a straight forward to recover the weights for the particular words comprising a topic, allowing for easy interpretation of an event's contents (e.g. Fig.~\ref{fig:topics}). Other models such as Google's universal sentence encoder offer a context-sensitive encoding of text passages, but the encoding space is complex and non-linear and thus, recovering the original words used to fit the model is not straight forward.

However, it's worth pointing out that our framework is divorced from the particular choice of language model. Moreover, many of the aspects of our framework could be swapped out for other choices. For example, the language model, the timeseries segmentation model and the video-recall matching function could all be customized for the particular problem. Indeed for some problems, recovery of the particular recall words may not be necessary, and thus other text-modeling approaches (such as universal sentence encoder) may be preferable. Future work will explore the influence of particular model choices on the framework's accuracy.

@@

\subsection*{Naturalistic extensions of classic list-learning analyses}
In traditional list-learning experiments, participants view a list of items (e.g., words) and then recall the items later.  Our video-recall event matching approach affords us the ability to analyze memory in a similar way. The video and recall events can be treated analogously to studied and recalled ``items'' in a list-learning study.  We can then extend classic analyses of memory performance and dynamics (originally designed for list-learning experiments) to the more naturalistic video recall task used in our study.

@@

>>>>

%\title{How is experience transformed into memory?}
%\title{A content-based framework for capturing stimulus and memory dynamics reveals event-like structure in naturalistic episodic recall}
%\title{A content-based model characterizing/describing/representing how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall}
%\title{A content-based framework describing the temporal dynamics of experiences and memories reveals event-like structure in naturalistic episodic recall}
%\title{A content-based model of how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall}
\title{A novel framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall}
%\title{Recalling naturalistic experiences exhibits event-like structure}

@@

\begin{abstract}
Our life experiences unfold over time and the temporal dynamics of their contents form unique \textit{experience trajectories}.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  We apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We find that the video and subsequent recall share an event-like structure, and that the ``shapes'' of the trajectories formed by participants' recountings were all highly similar to that of the original video. Further, participants' that recounted the events with a high level of precision and in a distinctive way had better subsequent memory performance. Lastly, we identified a network of brain structures that are sensitive to the shapes of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences. These results highlight the rich event-like structure of memories for our life experiences, and introduce a novel framework for mapping between dynamic life experiences and their mnemonic counterparts.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our memories are continuous, rather than discrete: removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, counting the number or proportion of precise recalls is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or shape) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener.

@@

Here, we introduce a novel content-based framework for linking dynamic experiences to memories. We hypothesized that the ``shape'' and the event-like structure of an experience would be preserved in memory. Further, we tested  1) whether the \textit{precision} with which the participant recounts each event and 2) how \textit{distinctive} each recall event is (relative to the other recalled events) relates to overall memory performance. Last, we identify networks of  brain structures whose responses (as participants watched the episode) reflected the shape of the episode, and how participants would later recount the episode.

\section*{Results}
To characterize the shape of the \textit{Sherlock} episode participants watched and their subsequent recountings of the episode, we used a topic model~\citep{BleiEtal03} to discover the latent thematic content in the video.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents; they return as output two matrices.  The first output is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes which mixture of topics is reflected in each document.

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed 2 novel metrics for quantifying naturalistic memory representations: precision and distinctiveness.  We define precision as the average correlation between each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that across participants, a higher precision score is correlated to both hand annotated memory performance (Pearson's $r(15) = 0.6, p = 0.011$) as well as the number of recall events estimated by our model (Pearson's $r(15) = 0.64, p = 0.005$). A second novel metric we introduce here is distinctiveness, or how unique the recall description was to each video event. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  Similar to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.83, p < 0.001$ and model derived memory: Pearson's $r(15) = 0.71, p = 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.

@@

The prior analyses leverage the correspondence between the 100-dimensional topic proportion matrices for the video and participants' recalls to characterize recall.  However, it is difficult to gain deep insights into that content solely by examining the topic proportion matrices (e.g., Figs.~\ref{fig:model}A, D) or the corresponding correlation matrices (Figs.~\ref{fig:model}B, E, \corrmats).  To visualize the time-varying high-dimensional content in a more intuitive way~\citep{HeusEtal18a} we projected the topic proportions matrices onto a two-dimensional space using Uniform Manifold Approximation and Projection~\citep[UMAP; ][]{McInHeal18}.  In this lower-dimensional space, each point represents a single video or recall event, and the distances between the points reflect the distances between the events' associated topic vectors (Fig.~\ref{fig:trajectory}). In other words, events that are near to each other in this space are more semantically similar.

@@

When modeling memory experiments, often times events (or items) and their later memories are treated as binary (or categorical in the case of confidence ratings) and independent events. Our novel framework allows one to assess memory performance in a more continuous way (e.g. precision), as well as analyze the correlational structure of each encoding event to each memory event (e.g. distinctiveness). Further and importantly, it allows for consideration of the actual content of the experience/memories, which is not typically modeled. Leveraging this, using 2 novel memory metrics we find that the successful memory performance is related to 1) the \textit{precision} with which the participant recounts each event and 2) how \textit{distinctive} each recall event is (relative to the other recalled events). The first finding suggests to us that the accuracy of recall for \textit{any individual event} may predict the overall amount of information recovered by the participant.  The second finding suggests that remembering/describing events in a unique way (relative to other recalled events) is also related to the quantity of content recovered. Intriguingly, prior studies show that pattern separation, or the ability to discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging \citep{StarEtal10, YassEtal11c, YassStar11b}.  Future work might explore how/whether the novel metrics introduce here compare between cognitively impoverished groups and healthy controls.

% NOTE: do you know of a good reference for WAS and word2vec ?
While a large number of language models exist (e.g. WAS, LSA, word2vec, universal sentence encoder) \citep{LandEtal98, CerEtal18}, here we use topic models for a few reasons.  First, topic models capture the \textit{essence} of a text passage devoid of the specific set and order of words used.  This was an important feature of our model since different people may accurately recall a scene using very different language. Secondly, words can mean different things in different contexts (e.g. baseball bat vs. the animal bat), and topic models are robust to this since words can be part of multiple topics.  Lastly, topic models provide a straight forward to recover the weights for the particular words comprising a topic, allowing for easy interpretation of an event's contents (e.g. Fig.~\ref{fig:topics}). Other models such as Google's universal sentence encoder offer a context-sensitive encoding of text passages, but the encoding space is complex and non-linear and thus, recovering the original words used to fit the model is not straight forward. However, it's worth pointing out that our framework is divorced from the particular choice of language model. Moreover, many of the aspects of our framework could be swapped out for other choices. For example, the language model, the timeseries segmentation model and the video-recall matching function could all be customized for the particular problem. Indeed for some problems, recovery of the particular recall words may not be necessary, and thus other text-modeling approaches (such as universal sentence encoder) may be preferable. Future work will explore the influence of particular model choices on the framework's accuracy.

@@

\subsubsection*{Naturalistic extensions of classic list-learning analyses}
In traditional list-learning experiments, participants view a list of items (e.g., words) and then recall the items later.  Our video-recall event matching approach affords us the ability to analyze memory in a similar way. The video and recall events can be treated analogously to studied and recalled ``items'' in a list-learning study.  We can then extend classic analyses of memory performance and dynamics (originally designed for list-learning experiments) to the more naturalistic video recall task used in our study.

@@

\subsubsection*{Novel naturalistic memory metrics}
